Parsing Postfix Log Files

### 2009-04-28 doc/logparser-acronyms.tex
Changes made after proofreading the preface and chapters 1-3.
Changes made after proofreading chapter 7 and the appendices.

\newglossaryentry{LMTP}{
    type=\acronymtype,
    name={LMTP},
    description={Local Mail Transfer Protocol},
    first={Local Mail Transfer Protocol (LMTP)}
}
\newglossaryentry{LMTP glossary}{
    name={Local Mail Transfer Protocol},
    description={
        \acronym{LMTP} is a protocol derived from \acronym{SMTP} that
        removes the need for the server to maintain a mail delivery queue,
        instead relying on the client to maintain it.  Typically the client
        is an \acronym{MTA}, and the server a delivery agent or a mail
        store.  Full details are available in~\cite{lmtp-rfc}.
    }
}

\newglossaryentry{MTA}{
    type=\acronymtype,
    name={MTA},
    description={Mail Transfer Agent},
    first={Mail Transfer Agent (MTA)}
}
\newglossaryentry{MTA glossary}{
    name={Mail Transfer Agent},
    description={
        \glsreset{MTA}A \acronym{MTA} sends and receives mail via
        \acronym{SMTP}\@.  Users submit mail to an \acronym{MTA} via their
        mail client (e.g.\ Microsoft Outlook, Thunderbird, webmail
        services); the sending \acronym{MTA} transfers the mail to the
        receiving \acronym{MTA}.
    }
}

@@

\newglossaryentry{pid}{
    type=\acronymtype,
    name={pid},
    description={Process Identifier},
    first={Process Identifier (pid)}
}
\newglossaryentry{pid glossary}{
    name={Process Identifier},
    description={
        There may be multiple copies of any program executing at any one
        time, so the program's name is not suitable as a distinguishing
        identifier; instead, each process is given a \acronym{pid} that is
        guaranteed to be unique for the lifetime of the process.  Once the
        process has completed, the pid may be reused, because they are
        drawn from a finite pool.
    }
}

@@

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A \acronym{CFG} is a grammar in which every transition or
        production takes the form $L\rightarrow{}R$, where $L$ is a single
        non-terminal symbol, and $R$ is a (possibly empty) string of
        terminal or non-terminal symbols.  Every \acronym{CFG} can be
        recognised by a \acronym{PDA}, and \acronyms{PDA} cannot recognise
        more complicated grammars, so \acronyms{CFG} and \acronyms{PDA} are
        equivalent.
    }
}

\newglossaryentry{PDA}{
    type=\acronymtype,
    name={PDA},
    description={Push-Down Automata},
    first={Push-Down Automata (PDA)}
}
\newglossaryentry{PDA glossary}{
    name={Push-Down Automata},
    description={
        \glsreset{PDA}A \acronym{PDA} is a computational device similar to
        a \acronym{FA}, but it can additionally use a stack to store data.
        A \acronym{PDA} can manipulate the stack during state transitions
        by adding or removing a single piece of data, and when determining
        which state transition to take, the piece of data on the top of the
        stack can be used in addition to the input.  In addition to the
        languages recognised by \acronyms{FA}, \acronyms{PDA} can recognise
        languages of the form $a^{n}b^{n}$, i.e.\ any number of one symbol
        followed by the same number of another symbol.
    }
}

>>>>

\newglossaryentry{LMTP}{
    type=\acronymtype,
    name={LMTP},
    description={Local Mail Transfer Protocol},
    first={Local Mail Transfer Protocol (LMTP)}
}
\newglossaryentry{LMTP glossary}{
    name={Local Mail Transfer Protocol},
    description={
        \acronym{LMTP} is a protocol derived from \acronym{SMTP} that
        removes the need for the server to maintain a mail delivery queue,
        instead relying on the client to maintain it.  Typically the client
        is an \acronym{MTA}, and the server is a delivery agent or a mail
        store.  Full details are available in~\cite{lmtp-rfc}.
    }
}

\newglossaryentry{MTA}{
    type=\acronymtype,
    name={MTA},
    description={Mail Transfer Agent},
    first={Mail Transfer Agent (MTA)}
}
\newglossaryentry{MTA glossary}{
    name={Mail Transfer Agent},
    description={
        \glsreset{MTA}A \acronym{MTA} sends and receives mail via
        \acronym{SMTP}\@.  Users submit mail to an \acronym{MTA} via their
        mail client (e.g.\ Microsoft Outlook, Thunderbird, webmail
        services); the sending \acronym{MTA} transfers the mail to the
        receiving \acronym{MTA}, which forwards the mail to another
        recipient, or delivers the mail to a user's mailbox or a program
        such as a mailing list manager.
    }
}

@@

\newglossaryentry{pid}{
    type=\acronymtype,
    name={pid},
    description={Process Identifier},
    first={Process Identifier (pid)}
}
\newglossaryentry{pid glossary}{
    name={Process Identifier},
    description={
        There may be multiple copies of any program executing at any one
        time, so the program's name is not suitable as a distinguishing
        identifier; instead, each process is given a \acronym{pid} that is
        guaranteed to be unique for the lifetime of the process.  Once the
        process has completed, the \acronym{pid} may be reused, because
        they are drawn from a finite pool.
    }
}

@@

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A \acronym{CFG} is a grammar in which every transition or
        production takes the form $L\rightarrow{}R$, where $L$ is a single
        non-terminal symbol, and $R$ is a (possibly empty) string of
        terminal or non-terminal symbols.  Every \acronym{CFG} can be
        recognised by a \acronym{PDA}, and \acronyms{PDA} cannot recognise
        more complicated grammars, so \acronyms{CFG} and \acronyms{PDA} are
        equivalent in computational power.
    }
}

\newglossaryentry{PDA}{
    type=\acronymtype,
    name={PDA},
    description={Push-Down Automata},
    first={Push-Down Automata (PDA)}
}
\newglossaryentry{PDA glossary}{
    name={Push-Down Automata},
    description={
        \glsreset{PDA}A \acronym{PDA} is a computational device similar to
        a \acronym{FA}, but it can additionally use a stack to store data.
        A \acronym{PDA} can manipulate the stack during state transitions
        by adding or removing a single piece of data, and when determining
        which state transition to take, the piece of data on the top of the
        stack can be used in addition to the input.  \acronym{PDA} can
        recognise all the languages recognised by \acronym{FA}, and can
        also recognised languages of the form $a^{n}b^{n}$.
    }
}

### 2009-04-28 doc/logparser-conclusion.tex
Changes made after proofreading chapter 7 and the appendices.

This architecture's greatest strength is the ease with which parsers based
on it can be adapted to deal with new requirements and inputs.  Parsing a
variation of an existing input is a trivial task: simply modify an existing
rule or add a new rule and the task is complete.  Parsing a new category of
input is achieved by writing a new action and a rule for each input
variant; quite often the new action will not need to interact with existing
actions, but when interaction is required the framework provides the
necessary facilities.  The architecture imposes very little red tape when
writing new actions, allowing the implementer to focus their time and
energy on correctly implementing their new action (\sectionref{actions in
architecture}).  The separation of the architecture into rules, actions,
and framework (\sectionref{parser design}) is unusual, partly because the
three are separated so completely.  Although parsers are often divided into
separate source code files (the combination of lex \&
yacc~\cite{lex-and-yacc-book} being a common example), the parts are
usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  This separation
enables the optimisations discussed in \sectionref{parser efficiency}, and
it also allows different approaches to ruleset management, e.g.\ using
machine learning techniques to seamlessly create or alter rules to
recognise new inputs (\sectionref{adding new rules in architecture}).  The
decoupling of rules from actions allows different sets of rules to be used
with one set of actions, e.g.\ a parser might have actions to process
versions one and two of a file format; by choosing the appropriate ruleset
the parser will parse version one, or version two, or both versions.  A
general purpose framework can be written, so that writing a parser just
requires writing actions and rules.  The architecture makes it possible to
apply commonly used programming techniques (such as object orientation,
inheritance, composition, delegation, roles, modularisation, or closures)
when designing and implementing a parser, simplifying the process of
working within a team or developing and testing additional functionality.
This architecture is ideally suited to parsing inputs that are not fully
understood or do not follow a fixed grammar: the architecture warns about
unrecognised inputs and errors encountered by actions, but continues
parsing as best it can, allowing the developer of a new parser to decide
which deficiencies are most important and require attention first, rather
than being forced to fix the first error that arises.

@@

When writing \parsername{}, the real difficulties arose once the parser was
successfully recognising almost all of the log lines, because most of the
irregularities and complications explained in \sectionref{complications}
started to become apparent then.  Adding new rules to deal with numerous
infrequently occurring log line variants was a simple if tiresome task,
whereas dealing with mails that were missing information or where Postfix's
actions were not being correctly reconstructed was much more grueling.
Trawling through log files was extremely time consuming and quite error
prone, searching for something out of the ordinary that might help diagnose
the problem, and eventually finding it --- sometimes hundreds or even
thousands of log lines away from the last occurrence of the queueid for the
mail in question.  Sometimes the task was not to identify the unusual log
line, but to spot that a log line normally present was missing, i.e.\ to
realise that one log line amongst thousands was absent.  In all cases the
evidence was used to construct a hypothesis to explain the irregularities,
and that hypothesis was tested in \parsername{}; if successful, the parser
was modified to deal with the irregularities, without adversely affecting
existing parsing.  The complications documented in
\sectionref{complications} are presented in the order they were solved in,
and that order closely resembles the frequency in which they occur; the
most frequently occurring complications dominate the warning messages
produced, and so naturally they were the first complications to be dealt
with.

@@

The data gathered by \parsername{} provides the foundation for the future
of this project: using machine-learning algorithms to analyse the data and
optimise the set of anti-spam defences in use, followed by identifying
patterns in the data that could be used to write new anti-spam techniques
to recognise and reject spam rather than accepting it.  The database
provides the data in a normalised form that is far easier to use as input
to new or existing implementations of machine-learning algorithms than
trying to adapt each algorithm to extract data directly from log files.
New policy servers, written to implement new anti-spam measures, can be
tested or trained by using the collected data to simulate mail delivery
attempts; this would allow simple, fast, reproducible testing, without the
risk of adversely affecting a production mail server.  Development of
\parsername{} is finished, i.e.\ it correctly parses Postfix log files, and
in future it will only require maintenance; however, one avenue of future
development under consideration is to extend it to parse non-Postfix log
lines, e.g.\ SpamAssassin or Amavisd-new log lines.  \parsername{} can
easily be extended to do this, but it requires a method of associating the
non-Postfix log lines with the existing data structures and state tables,
so that all of the data for a mail delivery attempt can be stored together.

>>>>

This architecture's greatest strength is the ease with which parsers based
on it can be adapted to deal with new requirements and inputs.  Parsing a
variation of an existing input is a trivial task: simply modify an existing
rule or add a new rule and the task is complete.  Parsing a new category of
input is achieved by writing a new action and a rule for each input
variant; quite often the new action will not need to interact with existing
actions, but when interaction is required the framework provides the
necessary facilities.  The architecture imposes very little red tape when
writing new actions, allowing the implementer to focus their time and
energy on correctly implementing their new action (\sectionref{actions in
architecture}).  The separation of the architecture into rules, actions,
and framework (\sectionref{parser design}) is unusual, partly because the
three are separated so completely.  Although parsers are often divided into
separate source code files (the combination of lex \&
yacc~\cite{lex-and-yacc-book} being a common example), the parts are
usually quite internally interdependent, and will be combined by the
compilation process; in contrast, \parsername{} keeps the rules and actions
separate until the parser runs (\sectionref{framework in implementation}).
This separation enables the optimisations discussed in \sectionref{parser
efficiency}, and it also allows different approaches to ruleset management,
e.g.\ using machine learning techniques to seamlessly create or alter rules
to recognise new inputs (\sectionref{adding new rules in architecture}).
The decoupling of rules from actions allows different sets of rules to be
used with one set of actions, e.g.\ a parser might have actions to process
versions one and two of a file format; by choosing the appropriate ruleset
the parser will parse version one, or version two, or both versions.  A
general purpose framework can be written, so that writing a parser just
requires writing actions and rules.  The architecture makes it possible to
apply commonly used programming techniques (such as object orientation,
inheritance, composition, delegation, roles, modularisation, or closures)
when designing and implementing a parser, simplifying the process of
working within a team or developing and testing additional functionality.
This architecture is ideally suited to parsing inputs that are not fully
understood or do not follow a fixed grammar: the architecture warns about
unrecognised inputs and errors encountered by actions, but continues
parsing as best it can, allowing the developer of a new parser to decide
which deficiencies are most important and require attention first, rather
than being forced to fix the first error that arises.

@@

When writing \parsername{}, the real difficulties arose once the parser was
successfully recognising almost all of the log lines, because most of the
irregularities and complications documented in \sectionref{complications}
started to become apparent then.  Adding new rules to deal with numerous
infrequently occurring log line variants was a simple if tiresome task,
whereas dealing with mails that were missing information or where Postfix's
actions were not being correctly reconstructed was much more grueling.
Trawling through log files was extremely time consuming and quite error
prone, searching for something out of the ordinary that might help diagnose
the problem, and eventually finding it --- sometimes hundreds or even
thousands of log lines away from the last occurrence of the queueid for the
mail in question.  Sometimes the task was not to identify the unusual log
line, but to spot that a log line normally present was missing, i.e.\ to
realise that one log line amongst thousands was absent.  In all cases the
evidence was used to construct a hypothesis to explain the irregularities,
and that hypothesis was tested in \parsername{}; if successful, the parser
was modified to deal with the irregularities, without adversely affecting
existing parsing.  The complications documented in
\sectionref{complications} are presented in the order they were solved in,
and that order closely resembles the frequency in which they occur; the
most frequently occurring complications dominated the warning messages
produced, and so naturally they were the first complications to be dealt
with.

@@

The data gathered by \parsername{} provides the foundation for the future
of this project: using machine-learning algorithms to analyse the data and
optimise the set of anti-spam defences in use, followed by identifying
patterns in the data that could be used to write new anti-spam techniques
to recognise and reject spam rather than accepting it.  The database
(\sectionref{database}) provides the data in a normalised form that is far
easier to use as input to new or existing implementations of
machine-learning algorithms than trying to adapt each algorithm to extract
data directly from log files.  New policy servers, written to implement new
anti-spam measures, can be tested or trained by using the collected data to
simulate mail delivery attempts; this would allow simple, fast,
reproducible testing, without the risk of adversely affecting a production
mail server.  Development of \parsername{} is finished, i.e.\ it correctly
parses Postfix log files, and in future it will only require maintenance;
however, one avenue of future development under consideration is to extend
it to parse non-Postfix log lines, e.g.\ SpamAssassin or Amavisd-new log
lines.  \parsername{} can easily be extended to do this, but it requires a
method of associating the non-Postfix log lines with the existing data
structures and state tables, so that all of the data for a mail delivery
attempt can be stored together.

### 2009-04-28 doc/logparser-glossary.tex
Changes made after proofreading chapter 7 and the appendices.

\newglossaryentry{Backscatter}{name={Backscatter},description={
    When a spam sender or virus sends mail with forged sender addresses,
    innocent servers are flooded with undeliverable mail notifications;
    this is called backscatter.
}}

@@

\newglossaryentry{<>}{name={$<>$},sort={<>},description={
    $<>$ is the sender address used for sending bounce notifications.  In
    the \acronym{SMTP} conversation all addresses are enclosed in $<>$, so
    \texttt{username@domain} is sent as \texttt{$<$username@domain$>$};
    thus $<>$ is actually an empty address, but it is always written as
    $<>$ for clarity.  All mail servers must accept mail sent from $<>$, or
    they are in violation of \acronym{RFC}~2821~\cite{RFC2821}.
}}

@@

\newglossaryentry{hash}{name={hash},description={
    A hashing function transforms a string of characters to a number.  A
    common usage is to maintain a data structure indexed by strings in an
    efficient manner.  A full description is beyond the scope of this
    thesis, further information can be found at~\cite{hash-functions}.
}}

\newglossaryentry{awk}{name={AWK},description={
    AWK is a general purpose programming language that is designed for
    processing text and is available as a standard utility on all Unix
    systems.
}}

\newglossaryentry{syslog}{name={Syslog},description={
    Syslog is the standard logging mechanism on Unix systems: a program
    sends log messages to syslog, then syslog filters and stores the
    messages as configured by the administrator.
}}

>>>>

\newglossaryentry{Backscatter}{name={Backscatter},description={
    When a spam sender or virus sends mail with forged sender addresses,
    innocent mail servers are flooded with undeliverable mail notifications
    from badly configured mail servers that either do not verify recipient
    addresses before accepting deliver attempts, or automatically reply to
    mail identified as spam; this is called backscatter.
}}

@@

\newglossaryentry{<>}{name={$<>$},sort={<>},description={
    $<>$ is the sender address used for sending bounce notifications.  In
    the \acronym{SMTP} conversation, all addresses are enclosed in $<>$, so
    \texttt{username@domain} is sent as \texttt{$<$username@domain$>$};
    thus $<>$ is actually an empty address, but it is always written as
    $<>$ for clarity.  Mail servers must not unconditionally reject mail
    sent from $<>$, or they are in violation of
    \acronym{RFC}~2821~\cite{RFC2821}.
}}

@@

\newglossaryentry{hash}{name={hash},description={
    A hashing function transforms a string of characters to a number.  A
    common usage is to maintain a data structure indexed by strings in an
    efficient manner.  A full description is beyond the scope of this
    thesis, further information can be found in~\cite{hash-functions}.
}}

\newglossaryentry{awk}{name={AWK},description={
    AWK is a general purpose programming language designed for processing
    text that is available as a standard utility on all Unix systems.
}}

\newglossaryentry{syslog}{name={Syslog},description={
    Syslog is the standard logging mechanism used on Unix systems: a
    program sends log messages to syslog, then syslog filters and stores
    the messages as configured by the administrator.
}}

### 2009-04-28 doc/logparser-postfix-daemons.tex
Changes made after proofreading chapter 7 and the appendices.

\newglossaryentry{postdrop}{name={postdrop},type={postfix},description={
    Postdrop is used when submitting mail locally on the server: it copies
    its input into a newly created mail in the queue, for processing by
    \daemon{pickup} and subsequent delivery.
    \daemonDocURL[\newline{}]{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
}}

\newglossaryentry{postsuper}{name={postsuper},type={postfix},description={
    Used by the administrator for maintenance tasks such as deleting mails
    from the queue, putting mail on hold and later releasing it, and
    consistency checking of the mail queue.
    \daemonDocURL[\newline{}]{http://www.postfix.org/postsuper.1.html}{2009/02/23}.
}}

@@

\newglossaryentry{sendmail}{name={sendmail},type={postfix},description={
    A Postfix component that is compatible with the Sendmail mail
    submission program which all Unix commands that need to send mail use;
    it executes \daemon{postdrop} to place a new mail in the queue.
    \daemonDocURL[\newline{}]{http://www.postfix.org/sendmail.1.html}{2009/02/23}.
}}

>>>>

\newglossaryentry{postdrop}{name={postdrop},type={postfix},description={
    Postdrop is used when submitting mail locally on the server: it copies
    its input into a newly created mail in the queue, for processing by
    \daemon{pickup} and subsequent delivery.
    \daemonDocURL{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
}}

\newglossaryentry{postsuper}{name={postsuper},type={postfix},description={
    Used by the administrator for maintenance tasks such as deleting mails
    from the queue, putting mail on hold and later releasing it, and
    consistency checking of the mail queue.
    \daemonDocURL{http://www.postfix.org/postsuper.1.html}{2009/02/23}.
}}

@@

\newglossaryentry{sendmail}{name={sendmail},type={postfix},description={
    A Postfix component that is compatible with the Sendmail mail
    submission program which all Unix commands that need to send mail use;
    it executes \daemon{postdrop} to place a new mail in the queue.
    \daemonDocURL{http://www.postfix.org/sendmail.1.html}{2009/02/23}.
}}
